---
title: "New York Accidents"
author: "Menna Mustafa"
date: "March 31, 2018"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

#Is this place safe to drive to? 

##This project aims to analyze New york accidents data 2016 and give score for unique 36,654 places according to how safe it is based on the answers of 3 questions
##1- How many accidents occur in that place?
##2- In how much time did they occur?
##3-How many people injured?

###Data source: https://www.kaggle.com/oscarleo/new-york-city-taxi-with-osrm/data

##Loading needed libraries
```{r echo=FALSE}
library(tidyverse)
library(lubridate)
library(leaflet)
library(janitor)
library(stringr)
library(ggthemes)
library(ggmap)
library(scales)
library(grid)
library(gridExtra)
library(NbClust)
library(HSAUR)
library(cluster)
```

##Dataset variables
```{r echo=FALSE}
names(accidents_2016)
```

>19.8% have NA location

###Making separate dataframe to exclude accidents with NA locations
```{r}
non_na_location_accidents <- accidents_2016 %>% 
  filter(!is.na(LOCATION))
```

#Q1:Where acciendent occur mostly?
```{r}
tabyl(non_na_location_accidents$LOCATION) %>% arrange(desc(n)) %>% head(n=10)
```

##For 36,654 location
```{r}
summary(tabyl(non_na_location_accidents$LOCATION)$n)
```

##Number of accidents distribution
```{r}
tabyl(non_na_location_accidents$LOCATION) %>% 
  ggplot(aes(x=n))+
  geom_histogram()
```

>Exponentially distributed so we can try make the score according to the IQR

```{r}
risk_scored_locations <-  non_na_location_accidents %>% 
  group_by(LOCATION) %>% 
  summarise(n = n()) %>%
  mutate(iqr = IQR(n),
            q1 = quantile(n, probs = 0.25),
            q3 = quantile(n, probs = 0.75)) %>% 
  ungroup() %>%
  mutate(
    risk_score = if_else(n < q1-(1.5*iqr) | n > q3+(1.5*iqr), 3,
                 if_else(n < q1-(1*iqr) | n > q3+(1*iqr),2,
                 if_else(n < q1-(.5*iqr) | n > q3+(.5*iqr),1,0)))
    )
```


##Distribution of risky places
```{r}
tabyl(risk_scored_locations$risk_score) 
```

>It seems like we have some sort of unbalanced categories

```{r}
risk_scored_locations %>%
  filter(risk_score==0) %>% tabyl(n)

risk_scored_locations %>%
  filter(risk_score==1) %>%
  tabyl(n)

risk_scored_locations %>%
  filter(risk_score==3) %>%
  tabyl(n)
```

###So, from 1 to 4 accidents get score 0 -> Zero risk place
### 5 accidents -> low risk
### 6 accidents -> medium risk
### 7 and above -> high risk 

##Think of another idea to categorize instead of IQR

#Logical method
###So, from 1 to 4 accidents -> Zero risk place
### 5 to 11 accidents -> low risk
### 12 to 24 accidents -> medium risk
### 45 to 63 -> high risk 

```{r}
locations_accidents <- non_na_location_accidents %>% 
  group_by(LOCATION) %>% 
  summarise(No_of_accidents = n())

locations_accidents$No_of_accidents_score <-
  ifelse(locations_accidents$No_of_accidents >0 & locations_accidents$No_of_accidents<=4,1,
  ifelse(locations_accidents$No_of_accidents >4 & locations_accidents$No_of_accidents<=11,2,
  ifelse(locations_accidents$No_of_accidents >11 & locations_accidents$No_of_accidents<=24,3,4)))
```

```{r}
table(locations_accidents$No_of_accidents_score)
```

> 1: No risk
  2: low risk
  3: Medium risk
  4: High risk

>Better Categories

##Manipulation for location accidents
```{r}
locations_accidents$LOCATION <- 
  str_replace_all(locations_accidents$LOCATION," ","")

```


```{r}
x <- non_na_location_accidents %>% 
  group_by(LOCATION) %>% 
  summarise(
    number_of_injuries = sum(`NUMBER OF PERSONS INJURED`)
  )
  
locations_accidents <- separate(locations_accidents, LOCATION, into = c("Longitude","Latitude"), sep="\\,")

locations_accidents$Longitude <- str_replace_all(locations_accidents$Longitude, "\\(","")
locations_accidents$Latitude <- str_replace_all(locations_accidents$Latitude,"\\)","")

locations_accidents <- left_join(locations_accidents,x, by="LOCATION")
```

#Explor number of injuries
```{r}
tabyl(locations_accidents$number_of_injuries)

locations_accidents %>%
  filter(number_of_injuries > 0) %>% 
  ggplot(aes(x=number_of_injuries))+
  geom_density()
```

#To give sore we will try make it by IQR first
```{r}
Injury_scored_locations <-  non_na_location_accidents %>% 
  group_by(LOCATION) %>% 
  summarise(n = sum(`NUMBER OF PERSONS INJURED`)) %>%
  mutate(iqr = IQR(n),
            q1 = quantile(n, probs = 0.25),
            q3 = quantile(n, probs = 0.75)) %>% 
  ungroup() %>%
  mutate(
    risk_score = if_else(n < q1-(1.5*iqr) | n > q3+(1.5*iqr), 3,
                 if_else(n < q1-(1*iqr) | n > q3+(1*iqr),2,
                 if_else(n < q1-(.5*iqr) | n > q3+(.5*iqr),1,0)))
    )

tabyl(Injury_scored_locations$risk_score)
```

>Seems like this method doesn't work

##Try clustering
```{r}
Injury_scored_locations <- non_na_location_accidents %>% 
  group_by(LOCATION) %>% 
  summarise(n = sum(`NUMBER OF PERSONS INJURED`))

tabyl(Injury_scored_locations$n)

```

>1: No risk (0)
2: low risk (1)
3: Medium risk (2~5)
4: high risk (>5)

##Injuries scoring
```{r}
locations_accidents$No_of_injury_score <-
  ifelse(locations_accidents$number_of_injuries == 0,1,
  ifelse(locations_accidents$number_of_injuries == 1,2,
  ifelse(locations_accidents$number_of_injuries>1 & locations_accidents$number_of_injuries<=5,3,4)))

tabyl(locations_accidents$No_of_injury_score)
```

#Time frame
```{r}
y <- non_na_location_accidents %>% 
  group_by(LOCATION) %>% 
  summarise(
    number_of_accidents = n(),
    first_date = last(datetime),
    last_date = first(datetime)
    #interval = last_date - first_date,
  )
y$time.interval <- y$last_date - y$first_date
y$duration <- as.duration(y$time.interval)

y$duration <- as.character(y$duration)
y %>% arrange(desc(duration))


y$duration <- gsub(".*\\((.*)\\).*", "\\1", y$duration)
y$duration <- str_replace_all(y$duration, "~","")
y$duration <- str_replace_all(y$duration, " weeks","")
y <- y %>% 
  select(LOCATION, number_of_accidents, duration)

y$duration <- str_replace_all(y$duration, "s","")

duration_in_days <- y %>% 
  filter(str_detect(duration, "day")) %>% 
  select(number_of_accidents, duration)

duration_in_weeks <- y %>% 
  filter(str_detect(duration, "week")) %>% 
  select(number_of_accidents, duration)

duration_in_weeks$duration <- str_replace_all(duration_in_weeks$duration, " week","")
duration_in_weeks$duration <- str_replace_all(duration_in_weeks$duration, "s","")
duration_in_weeks$duration <- as.numeric(duration_in_weeks$duration)

duration_in_weeks %>% 
  ggplot(aes(x=duration))+
  geom_histogram()
```

```{r}
non_na_location_accidents %>% 
  filter(LOCATION=="(40.732384, -74.0103519)") %>% 
  select(datetime) %>% 
  arrange(datetime)
```

>Validated!

```{r}
time_in_weeks_score <-  duration_in_weeks %>% 
  group_by(round(duration,0)) %>% 
  summarise(n = n()) %>%
  mutate(iqr = IQR(n),
            q1 = quantile(n, probs = 0.25),
            q3 = quantile(n, probs = 0.75)) %>% 
  ungroup() %>%
  mutate(
    risk_score = if_else(n < q1-(1.5*iqr) | n > q3+(1.5*iqr), 4,
                 if_else(n < q1-(1*iqr) | n > q3+(1*iqr),3,
                 if_else(n < q1-(.5*iqr) | n > q3+(.5*iqr),2,1)))
    )

tabyl(time_in_weeks_score$risk_score)
```

>Unbalanced!

```{r}
time_in_weeks_score %>% 
  ggplot(aes(x=`round(duration, 0)`))+
  geom_histogram()+
  scale_x_continuous(breaks = seq(0,25,1))
```

```{r}
time_in_weeks_score$risk_score <-
  ifelse(time_in_weeks_score$`round(duration, 0)` <9 ,4,
  ifelse(time_in_weeks_score$`round(duration, 0)` <16,3,
  ifelse(time_in_weeks_score$`round(duration, 0)` <22,2,1)))

tabyl(time_in_weeks_score$risk_score)
```

##So, we can consider the following model for scoring time intervals
>If that number of accidents happen in less than 17 weeks -> high score
else -> lower score

```{r}
y <- y %>% 
  mutate(
    id  = seq(1,36654,1)
  )

duration_in_weeks <- y %>% 
  filter(str_detect(duration, "week") | str_detect(duration, "weeks")) %>% 
  select(id, number_of_accidents, duration)

##Cleaning
duration_in_weeks$duration <- gsub(".*\\((.*)\\).*", "\\1", duration_in_weeks$duration)
duration_in_weeks$duration <- str_replace_all(duration_in_weeks$duration, "~","")
duration_in_weeks$duration <- str_replace_all(duration_in_weeks$duration, " weeks","")
duration_in_weeks$duration <- str_replace_all(duration_in_weeks$duration, " week","")
duration_in_weeks$duration <- round(as.numeric(duration_in_weeks$duration),0)

duration_in_weeks <- duration_in_weeks %>% 
  mutate(
    time_score = ifelse(duration>15,1,2)
  )
tabyl(duration_in_weeks$time_score)
```

>If accidents happen in days, higher score

```{r}
duration_in_days <- y %>% 
  filter(str_detect(duration, "day")) %>% 
  select(id, number_of_accidents, duration)

duration_in_days$duration <- gsub(".*\\((.*)\\).*", "\\1", duration_in_days$duration)
duration_in_days$duration <- str_replace_all(duration_in_days$duration, "~","")
duration_in_days$duration <- str_replace_all(duration_in_days$duration, " days","")
duration_in_days$duration <- str_replace_all(duration_in_days$duration, " day","")
duration_in_days$duration <- round(as.numeric(duration_in_days$duration),0)

duration_in_days <- duration_in_days %>% 
  mutate(
    time_score = 3
  )
```

>If accidents happen in hours or minutes , much higher risk!

```{r}
duration_in_hours <- y %>% 
  filter(str_detect(duration, "hour") | str_detect(duration, "min")) %>% 
  select(id, number_of_accidents, duration)

duration_in_hours$duration <- gsub(".*\\((.*)\\).*", "\\1", duration_in_hours$duration)
duration_in_hours$duration <- str_replace_all(duration_in_hours$duration, "~","")
duration_in_hours$duration <- str_replace_all(duration_in_hours$duration, " hours","")
duration_in_hours$duration <- str_replace_all(duration_in_hours$duration, " hour","")
duration_in_hours$duration <- str_replace_all(duration_in_hours$duration, " minutes","")
duration_in_hours$duration <- str_replace_all(duration_in_hours$duration, " minute","")
duration_in_hours$duration <- round(as.numeric(duration_in_hours$duration),0)

duration_in_hours <- duration_in_hours %>% 
  mutate(
    time_score = 4
  )
```

```{r}
duration_0 <- y %>% 
  filter(duration=="0s") %>% 
  select(id, number_of_accidents, duration)

duration_0$duration <- str_replace_all(duration_0$duration, "s","")
duration_0$duration <- as.numeric(duration_0$duration)

duration_0 <- duration_0 %>% 
  mutate(
    time_score = ifelse(number_of_accidents>1,4,1)
  )
table(duration_0$number_of_accidents)
tabyl(duration_0$time_score)
```

```{r}
duration_score <- rbind(duration_in_days, duration_0, duration_in_weeks, duration_in_hours)
```

```{r}
names(duration_score)
duration_score <- duration_score %>% 
  select(id, time_score)

left_join(y,duration_score, by="id")

y <- left_join(y,duration_score, by="id") %>% 
  select(id, duration,  time_score)

locations_accidents <- locations_accidents %>% 
  mutate(
    id = seq(1,36654,1)
  )

locations_accidents <- left_join(locations_accidents, y, by="id")
```

##Now, We have 3 components that will make our final scoring
>Score based on number of accidents
>Score based on how many person got injuried
>Score based on time interval

```{r}
locations_accidents <-locations_accidents %>% 
  mutate(
    R_score = 0.5*time_score + 0.3*No_of_injury_score + 0.2*No_of_accidents_score
  )

tabyl(locations_accidents$R_score)

locations_accidents %>% 
  select(No_of_accidents, number_of_injuries, duration, R_score) %>% 
  View()
```

```{r}
locations_accidents %>% 
  ggplot(aes(x=R_score))+
  scale_x_log10()+
  geom_density()
```


##finaly clean locations for maps
```{r}
locations_accidents <- separate(locations_accidents, LOCATION, into = c("Longitude","Latitude"), sep="\\,")

locations_accidents$Longitude <- str_replace_all(locations_accidents$Longitude, "\\(","")
locations_accidents$Latitude <- str_replace_all(locations_accidents$Latitude,"\\)","")
```

#Let's visualize high risk places, as clusters of places/polygons
```{r}
getColor <- function(locations_accidents) {
  sapply(locations_accidents$No_of_accidents_score, function(No_of_accidents_score) {
  if(No_of_accidents_score == 4) {
    "red" }
   else if(No_of_accidents_score==3) {
    "orange"
     }
    else if(No_of_accidents_score==2) {
      "green"
    }})
}

icons <- awesomeIcons(
  icon = 'ios-close',
  iconColor = 'black',
  library = 'ion',
  markerColor = getColor(locations_accidents)
)

locations_accidents %>% filter(No_of_accidents_score>1) %>% 
  leaflet() %>% 
  addTiles() %>%
  addMarkers(~as.numeric(Latitude), ~as.numeric(Longitude), icon=icons, label = ~as.character(
    paste("Number of accidents: ",No_of_accidents,"\n Number of people injured: ",
          number_of_injuries,"\n Time interval: ",duration)
  ), clusterOptions = markerClusterOptions())
```