---
title: "Congestion Analysis"
author: "Menna Mustafa"
date: "April 22, 2018"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---
#This notebook will use new york taxi trips routes for congestion analysis to answer mainly 3 questions:

###1- What are the top 10 congested streets in NYC?
###2- When are they mostly congested? (Peak hour, peak day)
###3- Where is the peak point of pickups?

```{r setup, include<-FALSE, echo<-FALSE}
knitr::opts_chunk$set(echo <- TRUE,warning <- FALSE, message <- FALSE)
```

```{r echo=FALSE}
library(tidyverse)
library(lubridate)
library(data.table)
library(sp)
library(rgeos)
library(geosphere)
library(ggmap)
library(RColorBrewer)
library(treemap)
library(gganimate)
library(gridExtra)
library(leaflet)
```

##fastest routes are the whole routes that we have

```{r echo=FALSE}
##To get pickup datetime we are going to join with trips data
train <- train %>% 
  select(id, pickup_datetime, dropoff_datetime,trip_duration)

fastest_routes <- left_join(fastest_routes, train, by="id")

#Mutate new time variables 
fastest_routes$pickup_hour <- hour(fastest_routes$pickup_datetime)
fastest_routes$pickup_week <- week(fastest_routes$pickup_datetime)
fastest_routes$pickup_month <- month(fastest_routes$pickup_datetime)
fastest_routes$pickup_day <-wday(fastest_routes$pickup_datetime)
```


```{r echo=FALSE}
##Change day format
fastest_routes$pickup_day <- 
  ifelse(fastest_routes$pickup_day==1,"Sunday",
  ifelse(fastest_routes$pickup_day==2,"Monday",
  ifelse(fastest_routes$pickup_day==3,"Tuesday",
  ifelse(fastest_routes$pickup_day==4,"Wednesday",
  ifelse(fastest_routes$pickup_day==5,"Thursday",
  ifelse(fastest_routes$pickup_day==6, "Friday","Saturday"))))))
```

##Extract all streets from the dataset
```{r echo=FALSE}
steps_streets <- fastest_routes  %>% 
  select(id, street_for_each_step, pickup_day, pickup_hour, pickup_datetime, trip_duration)

streets_routes <- separate_rows(steps_streets, street_for_each_step,sep = "\\|", convert = TRUE)

streets_routes <- streets_routes %>% 
  group_by(id) %>% 
  mutate(
    street_time = pickup_datetime + (row_number() * trip_duration)/n()
  )

streets_routes$street_hour <- hour(streets_routes$street_time)
streets_routes$street_day <-wday(streets_routes$street_time)

streets_routes$street_day <- 
  ifelse(streets_routes$street_day==1,"Sunday",
  ifelse(streets_routes$street_day==2,"Monday",
  ifelse(streets_routes$street_day==3,"Tuesday",
  ifelse(streets_routes$street_day==4,"Wednesday",
  ifelse(streets_routes$street_day==5,"Thursday",
  ifelse(streets_routes$street_day==6, "Friday","Saturday"))))))
```

##Now we have each streets with number of occurences in all routes

##Lets get the top 10 streets
```{r echo=FALSE}
top10 <- streets_routes %>% 
  group_by(street_for_each_step) %>%
  summarise(
    number_of_routes_pass_street = n()
  ) %>% 
  arrange(desc(number_of_routes_pass_street)) %>% 
  head(n=10)
top10
```

##Top 10 streets that routes pass by analysis

###1- FDR Dive
```{r echo=FALSE}
names(streets_routes) <- c("id", "pickup_day", "pickup_hour", "pickup_datetime", 
                           "trip_duration", "street_for_each_step", "street_time",
                           "street_hour", "Day of the Week")
streets_routes  %>%
group_by(street_for_each_step, `Day of the Week`, street_hour) %>% 
summarise(
  number_of_routes = n()
) %>% 
   filter(street_for_each_step == "FDR Drive") %>%
  ggplot(aes(x=street_hour, y=number_of_routes, color=`Day of the Week`))+
   geom_line()+
xlab("Hour of the day")+
ylab("Number of trips passing through this street")+
ggtitle(paste("Congestion analysis for street: FDR Drive"))
  
```

###2-Broadway
```{r echo=FALSE}
streets_routes  %>%
group_by(street_for_each_step, `Day of the Week`, street_hour) %>% 
summarise(
  number_of_routes = n()
) %>% 
   filter(street_for_each_step == "Broadway") %>%
  ggplot(aes(x=street_hour, y=number_of_routes, color=`Day of the Week`))+
   geom_line()+
xlab("Hour of the day")+
ylab("Number of trips passing through this street")+
ggtitle(paste("Congestion analysis for street: Broadway"))
```

###3- 3rd Avenue
```{r echo=FALSE}
streets_routes  %>%
group_by(street_for_each_step, `Day of the Week`, street_hour) %>% 
summarise(
  number_of_routes = n()
) %>% 
   filter(street_for_each_step == "3rd Avenue") %>%
  ggplot(aes(x=street_hour, y=number_of_routes, color=`Day of the Week`))+
   geom_line()+
xlab("Hour of the day")+
ylab("Number of trips passing through this street")+
ggtitle(paste("Congestion analysis for street: 3rd Avenue"))
```

###4- 6th Avenue
```{r echo=FALSE}
streets_routes  %>%
group_by(street_for_each_step, `Day of the Week`, street_hour) %>% 
summarise(
  number_of_routes = n()
) %>% 
   filter(street_for_each_step == "6th Avenue") %>%
  ggplot(aes(x=street_hour, y=number_of_routes, color=`Day of the Week`))+
   geom_line()+
xlab("Hour of the day")+
ylab("Number of trips passing through this street")+
ggtitle(paste("Congestion analysis for street: 6th Avenue"))
```

###5-1st Avenue
```{r echo=FALSE}
streets_routes  %>%
group_by(street_for_each_step, `Day of the Week`, street_hour) %>% 
summarise(
  number_of_routes = n()
) %>% 
   filter(street_for_each_step == "1st Avenue") %>%
  ggplot(aes(x=street_hour, y=number_of_routes, color=`Day of the Week`))+
   geom_line()+
xlab("Hour of the day")+
ylab("Number of trips passing through this street")+
ggtitle(paste("Congestion analysis for street: 1st Avenue"))
```


##For all new york
```{r echo=FALSE}
streets_routes %>% 
  group_by(`Day of the Week`, street_hour) %>% 
  summarise(
    n=n()
  ) %>% 
  ggplot(aes(x=street_hour, y=n, color=`Day of the Week`))+
   geom_line()+
xlab("Hour of the day")+
ylab("Number of trips passing")+
ggtitle(paste("Congestion analysis for New York"))
```


##How many pickups are made from these streets?
```{r echo=FALSE}
fastest_routes %>% 
  group_by(starting_street) %>% 
  filter(starting_street %in% top10$street_for_each_step) %>% 
  summarise(
    number_of_pickups = n()
  ) %>% 
  arrange(desc(number_of_pickups))
```

###Map Visualization
```{r}
getColor <- function(locs) {
  sapply(locs$number_of_pickups, function(number_of_pickups) {
  if(number_of_pickups > 50000) {
    "red" }
   else if(number_of_pickups > 40000) {
    "orange"
     }
    else{
      "green"
    }})
}

icons <- awesomeIcons(
  icon = 'ios-close',
  iconColor = 'black',
  library = 'ion',
  markerColor = getColor(locs)
)

locs %>% 
  leaflet() %>% 
  addTiles() %>%
  addMarkers(~as.numeric(lat), ~as.numeric(lon), icon=icons, label = ~as.character(
    paste("Number of pickups for 2016: ",number_of_pickups)))
```

