---
title: "New York Taxi"
author: "Menna Mustafa"
date: "April 23, 2018"
output: html_document
---

##This Notebook is for analyze and visualize new york taxi trips dataset to predict trip duration 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Libraries needed
```{r}
library(tidyverse)
library(lubridate)
library(data.table)
library(sp)
library(rgeos)
library(geosphere)
library(ggmap)
library(RColorBrewer)
library(treemap)
library(gganimate)
library(gridExtra)
library(caret)
library(xgboost)
library(neuralnet)
```

##Structure and variables of our data
```{r}
glimpse(train)
```

##Explore our target variable

```{r}
train %>% 
  ggplot(aes(x=trip_duration)) + 
  geom_histogram(bins=40000, fill="blue")+
  theme_bw()+theme(axis.title = element_text(size=12),axis.text = element_text(size=12))+
  ylab("Density")+coord_cartesian(x=c(0,6000))
```

##Let's try to take log scale

```{r}
train %>% 
  ggplot(aes(x=log10(trip_duration)))+
  geom_density(fill="red", alpha=0.2)
```

>That will make our analysis much easier

##Take the log of trip duration
```{r}
train$trip_duration <- log10(train$trip_duration)
```

##Check locations for outliers
```{r}
train %>% 
  ggplot(aes(x=pickup_longitude))+
  geom_density()

train %>% 
  ggplot(aes(x=pickup_latitude))+
  geom_density()

train %>% 
  ggplot(aes(x=dropoff_longitude))+
  geom_density()

train %>% 
  ggplot(aes(x=dropoff_latitude))+
  geom_density()
```

>Observed outliers locations that make the distribution looks like that, so we will remove those outliers

```{r}
train <- train %>% 
  filter(
    (train$pickup_latitude > 40.6) & (train$pickup_latitude < 40.9) &
    (train$dropoff_latitude>40.6) & (train$dropoff_latitude < 40.9) &
    (train$dropoff_longitude > -74.05) & (train$dropoff_longitude < -73.7) &
    (train$pickup_longitude > -74.05) & (train$pickup_longitude < -73.7) 
  )
```

##Replot locations

```{r}
train %>% 
  ggplot(aes(x=pickup_longitude))+
  geom_density(fill="blue", alpha=0.2)

train %>% 
  ggplot(aes(x=pickup_latitude))+
  geom_density(fill="red", alpha=0.2)

train %>% 
  ggplot(aes(x=dropoff_longitude))+
  geom_density(fill="blue", alpha=0.2)

train %>% 
  ggplot(aes(x=dropoff_latitude))+
  geom_density(fill="red", alpha=0.2)
```

>Distribution becomes better after remove locations that make problem

##Join with fastest route data
```{r}
f_routes <- fastest_routes %>% 
  select(id, starting_street, end_street, street_for_each_step, total_distance)
train <- left_join(train, f_routes, by="id")
```

#Exploring distance
```{r}
train %>% 
  ggplot(aes(x=total_distance))+
  geom_density(fill="black", alpha=0.2)
```

```{r}
train %>% 
  filter(total_distance<20000) %>% 
  ggplot(aes(x=total_distance))+
  geom_density(fill="black", alpha=0.2)
```


```{r}
train %>% 
  ggplot(aes(x=total_distance))+
  scale_x_log10()+
  geom_density(fill="black", alpha=0.2)
```

##So we will take the log and remove outliers
```{r}
train %>% 
  filter(log10(total_distance)>2.5) %>% 
  ggplot(aes(x=log10(total_distance)))+
  geom_density(fill="black", alpha=0.2)
```

##To normalize distances, remove log10 distance <2.5 (14387)
```{r}
train <- train %>% 
  filter(log10(total_distance)>2.5)
```

```{r}
train$total_distance <- log10(train$total_distance)
```

##We have 2 vendors, let's explain them
```{r}
train$vendor_id <- factor(train$vendor_id)
train %>%
  ggplot(aes(x=vendor_id, y=trip_duration))+
  geom_boxplot()
```


#Extract some features from pickup datetime
```{r}
train$pickup_hour <- hour(train$pickup_datetime)
train$pickup_week <- week(train$pickup_datetime)
train$pickup_month <- month(train$pickup_datetime)
train$pickup_weekdays <- weekdays(train$pickup_datetime)
train$pickup_weekend <- ifelse(train$pickup_weekdays=="Thursday"|train$pickup_weekdays=="Friday","Weekend","not-Weekend")

train = as.data.table(train)

train[,pickup_datetime:=as.Date(pickup_datetime)]
train[,dropoff_datetime:=as.Date(dropoff_datetime)]

train[,":="(
   pickup_yday=yday(pickup_datetime)
  ,pickup_mday=mday(pickup_datetime)
)]
train <- as.data.frame(train)
```

##Calculate distance to validate that we have correct distance

```{r}
train <- as.data.table(train)
train <- train[,distance_Haversine := 
                     distHaversine(matrix(c(pickup_longitude, pickup_latitude), ncol = 2),
                     matrix(c(dropoff_longitude,dropoff_latitude), ncol = 2))
              ]
train <- as.data.frame(train)
```

##Convert also to log scale
```{r}
train %>% 
  ggplot(aes(x=log10(distance_Haversine)))+
  geom_density()

train <- train %>% 
  filter(log10(distance_Haversine)>2.5)

train$distance_Haversine <- log10(train$distance_Haversine)
```

##Explore relation between duration and vendors for each week day
```{r}
train$pickup_weekdays <- 
  ifelse(
    train$pickup_weekdays=="Saturday",0,
    ifelse(train$pickup_weekdays=="Sunday",1,
    ifelse(train$pickup_weekdays=="Monday",2,
    ifelse(train$pickup_weekdays=="Tuesday",3,
    ifelse(train$pickup_weekdays=="Wednesday",4,
    ifelse(train$pickup_weekdays=="Thursday",5,6))))))

train %>% 
 group_by(vendor_id, pickup_weekdays) %>% 
  summarise(
    mean_trip_duration = mean(10^(trip_duration))
  ) %>% 
  ggplot(aes(x=pickup_weekdays, y=mean_trip_duration, color=vendor_id))+
  geom_line()
```

>WOHOOO We got a feature! there is noticable difference between 2 vendors in trips duration

##Explore passenger count
```{r}
train %>% 
  group_by(passenger_count) %>% 
  summarise(
    mean_trip_duration = mean(10^(trip_duration))
  ) %>% 
  ggplot(aes(x=passenger_count, y=mean_trip_duration))+
  geom_line()

cor.test(train$passenger_count, train$trip_duration)
```

>By logic passenger count could affect trip duration, but .. data talks!

##Check days, trip duration and vendor
```{r}
##Check days
train$pickup_weekdays <- 
  ifelse(
    train$pickup_weekdays==0,"Saturday",
    ifelse(train$pickup_weekdays==1,"Sunday",
    ifelse(train$pickup_weekdays==2,"Monday",
    ifelse(train$pickup_weekdays==3,"Tuesday",
    ifelse(train$pickup_weekdays==4,"Wednesday",
    ifelse(train$pickup_weekdays==5,"Thursday","Friday"))))))

train %>% 
  ggplot(aes(x=pickup_weekdays, y=trip_duration, color=vendor_id))+
  geom_boxplot()
```

#Explore days with hours
```{r}
train$pickup_weekdays <- factor(train$pickup_weekdays)
train %>% 
  group_by(pickup_weekdays, pickup_hour) %>% 
  summarise(
    mean_trip_duration = mean(exp(trip_duration))
  ) %>% 
  ggplot(aes(x=pickup_hour, y=mean_trip_duration, color=pickup_weekdays))+
  geom_line()
```

> Which hour in day is also significant -> Got another feature!

##Explore store and forward flag
```{r}
train$store_and_fwd_flag <- 
  ifelse(train$store_and_fwd_flag=="N",0,1)
train$store_and_fwd_flag <- factor(train$store_and_fwd_flag)
train %>% 
  ggplot(aes(x=store_and_fwd_flag, y=10^(trip_duration)))+
  geom_boxplot()
```

>We have some outliers needs to be removed!

```{r}
train <- train %>% 
  filter(10^(trip_duration) < 1500000)
```

##Categorize trip durations and explore number of clients per category
```{r}

train %>%
  mutate(
      trip_duration_minutes = round(10^(trip_duration) / 60, 0),
      duration_grp = ifelse(trip_duration_minutes >=0 & trip_duration_minutes <=9, "Less than 10 minutes",ifelse(trip_duration_minutes >=10 & trip_duration_minutes <=29, "10 to 29 minutes", ifelse(trip_duration_minutes >=30 & trip_duration_minutes <=59, "30 to 59 minutes","60 or more minutes" ))),
    duration_grp = factor(duration_grp, 
                          levels = c("Less than 10 minutes", "10 to 29 minutes", "30 to 59 minutes", "60 or more minutes"))
  ) %>%
  group_by(duration_grp) %>% 
  summarise(passenger_count=sum(passenger_count, na.rm=TRUE)) %>% 
  ungroup %>%
  ggplot(aes(x=duration_grp, y=passenger_count)) + 
  geom_bar(stat="identity", fill="#E41A1C") +
  scale_y_continuous() +
  labs(x="Group", y="Passengers") +
  theme_minimal()+
  ggsave("groups.png")

```

>Most of our passengers are making trips that takes from 10~29 minutes

#Repeat previous plot but to compare vendors trips
```{r}
train %>%
  mutate(
      trip_duration_minutes = round(10^(trip_duration) / 60, 0),
      duration_grp = ifelse(trip_duration_minutes >=0 & trip_duration_minutes <=9, "Less than 10 minutes",ifelse(trip_duration_minutes >=10 & trip_duration_minutes <=29, "10 to 29 minutes", ifelse(trip_duration_minutes >=30 & trip_duration_minutes <=59, "30 to 59 minutes","60 or more minutes" ))),
    duration_grp = factor(duration_grp, 
                          levels = c("Less than 10 minutes", "10 to 29 minutes", "30 to 59 minutes", "60 or more minutes"))
  ) %>%
  group_by(duration_grp, vendor_id) %>% 
  summarise(passenger_count=sum(passenger_count, na.rm=TRUE)) %>% 
  ungroup %>%
  ggplot(aes(x=duration_grp, y=passenger_count, fill=vendor_id)) + 
  geom_bar(stat="identity") +
  scale_y_continuous() +
  labs(x="Group", y="Passengers") +
  theme_minimal()
```

> Vendor 2 takes most trips, that could indecate better service or more market share

##Trip Duration by Time of Day
```{r}
train %>%
  group_by(pickup_hour) %>% 
  summarise(trip_duration=median(10^(trip_duration), na.rm=TRUE)) %>% 
  ungroup %>%
  ggplot(aes(x=pickup_hour, y=trip_duration)) + 
  geom_bar(stat="identity", fill="yellow") +
  ylim(0, 1500) +
  labs(x="Hour", y="'Median of trips duration") +
  theme_minimal()
```

```{r}
train %>% 
  group_by(pickup_hour) %>% 
  summarise(median_trip_duration = median(10^(trip_duration))) %>% 
  ggplot(aes(x = pickup_hour, y = median_trip_duration)) + 
  geom_point(colour = "blue", size = 3) + 
  ggtitle("Taxi Trip duration per hour") + 
  xlab("Pickup Hour") + 
  ylab("Median trip time (seconds)")
```

>Pickup hour -> New feature

##Week days
```{r}
train$pickup_weekdays <- 
  ifelse(
    train$pickup_weekdays=="Saturday",0,
    ifelse(train$pickup_weekdays=="Sunday",1,
    ifelse(train$pickup_weekdays=="Monday",2,
    ifelse(train$pickup_weekdays=="Tuesday",3,
    ifelse(train$pickup_weekdays=="Wednesday",4,
    ifelse(train$pickup_weekdays=="Thursday",5,6))))))
train %>% 
  group_by(pickup_weekdays) %>% 
  summarise(median_trip_duration = median(trip_duration)) %>% 
  ggplot(aes(x = pickup_weekdays, y = median_trip_duration)) + 
  geom_point(colour = "blue", size = 3) + 
  ggtitle("NYC taxi Trip duration per day") + 
  xlab("Pickup day") + 
  ylab("Median trip time (seconds)")
```

>Day affects the duration of trips!

##Exlore on month level

```{r}
train %>% 
  group_by(pickup_month) %>% 
  summarise(median_trip_duration = median(trip_duration)) %>% 
  ggplot(aes(x = pickup_month, y = median_trip_duration)) + 
  geom_point(colour = "blue", size = 3) + 
  ggtitle("NYC taxi Trip duration per month") + 
  xlab("Pickup Month") + 
  ylab("Median trip time (seconds)")
```

>No conclusion can be done as the dataset doesn't contain any rides after June. but increase is noticed

##Save created variables
```{r}
train$trip_duration_minutes <- round(train$trip_duration/60,0)
train$duration_grp <-ifelse(train$trip_duration_minutes >=0 & train$trip_duration_minutes <=9, "Less than 10 minutes",ifelse(train$trip_duration_minutes >=10 & train$trip_duration_minutes <=29, "10 to 29 minutes", ifelse(train$trip_duration_minutes >=30 & train$trip_duration_minutes <=59, "30 to 59 minutes","60 or more minutes" )))
train$duration_grp = factor(train$duration_grp, 
                            levels = c("Less than 10 minutes", "10 to 29 minutes", "30 to 59 minutes", "60 or more minutes"))

```

##We want to think from vendors point of view, when most rides requested?
```{r}
##Explore number of rides per hour
train %>%
  group_by(pickup_hour) %>% 
  summarise(
    N= n()
  ) %>% 
  ggplot(aes(x=pickup_hour, y=N)) + 
  geom_bar(stat='identity', fill='steelblue') + 
  labs(x='', y='Number of Rides', title='Total Rides Per Hour')
```

>Increase after 7:00 AM --> People are going to schools, universties or work :D

##That could suppot congestion analysis

```{r}
##Speed of the ride per each hour in day
train$speed <- ((10^(train$total_distance))/1000) / ((10^(train$trip_duration))/3600)

train %>% 
  ggplot(aes(x=speed)) + 
  geom_histogram(bins=4000, fill="black")+
  theme_bw()+theme(axis.title = element_text(size=11),axis.text = element_text(size=8))+
  ylab("Density")+coord_cartesian(x=c(0,50))
```

```{r}
summary(train$speed)
```

>These numbers could indecate high congestion, let's explore more

```{r}
train %>% 
  group_by(pickup_hour) %>% 
  summarize(mean_speed = mean(speed),n()) %>% 
  ggplot(aes(x=pickup_hour,y=mean_speed))+
  geom_smooth(method = 'loess',color="grey10")+
  geom_point(color="red")+coord_cartesian(ylim=c(10,25))+theme_bw()
```

> Nearly around 7 AM ->> Congestion started and lower cars speed

```{r}
train %>% 
  group_by(pickup_hour) %>% 
  summarize(mean_trip_duration = mean(10^(trip_duration)),n()) %>% 
  ggplot(aes(x=pickup_hour,y=mean_trip_duration))+
  geom_smooth(method = 'loess',color="grey10")+
  geom_point(color="red")+coord_cartesian(ylim=c(500,1500))+theme_bw()
```

>Trip duration increases after 7 AM and reach the peak at 3:00 PM

##Let's look at distance
```{r}
##Distance in km
train$distance_km <- ((10^(train$total_distance))/1000)
##digging more
summary(train$distance_km)
```

#Can we highlight certain areas in New York that trips in it take longer time? 
```{r}
train_geo_data <- train %>% 
  group_by(round(pickup_longitude, 2),
           round(pickup_latitude, 2)) %>% 
  summarise(median_duration = median(10^(trip_duration)),
            median_distance = round(median(distance_km),1))


train_geo_data %>%
  ggplot()  + 
  geom_point(aes(x = `round(pickup_longitude, 2)`, y = `round(pickup_latitude, 2)`, colour = log(median_duration))) + 
  scale_x_continuous(limits = c(-74.4, -73.6)) + 
  scale_y_continuous(limits = c(40.4, 41)) +
  ggtitle("Time difference of taxi pick ups in NYC") + 
  xlab("Pick up longitude") + 
  ylab("Pick up latitude") + 
  scale_color_continuous(low = "blue", high = "#F5B317", name = "Log base 10 to median duration of trip")
```

>That's very important: So, If you want to take ride you will probably need to change your pickup location to avoid having longer trip duration time!

>Note that: This result different from the one in presentation as this one is after preprocessing

##Let's validate this point with numbers 
```{r}
train_geo_data %>% select(median_duration, median_distance) %>% 
  mutate(
    median_duration_log = log(median_duration)
  ) %>% 
  filter(median_duration_log>7.5) %>% 
  select(contains("median_dist")) %>% 
  summary()
```

> Numbers nearly confirmed our conclusion because distances are relatively small to take such long duration

##Let's do the same thing with drop_off locations
```{r}
train_geo_data_drop <- train %>% 
  group_by(round(dropoff_longitude, 2),
           round(dropoff_latitude, 2)) %>% 
  summarise(median_duration = median(10^(trip_duration)))

names(train_geo_data_drop) <- c("dropoff_lon", "dropoff_lat", "median_duration")

train_geo_data_drop %>%
  ggplot()  + 
  geom_point(aes(x = dropoff_lon, y = dropoff_lat, colour = log(median_duration))) + 
  scale_x_continuous(limits = c(-74.4, -73.6)) + 
  scale_y_continuous(limits = c(40.4, 41)) +
  ggtitle("Time difference of taxi drop offs in NYC") + 
  xlab("dropoff longitude") + 
  ylab("dropoff latitude") + 
  scale_color_continuous(low = "blue", high = "#F5B317", name = "Log base 10 to median duration of trip")
```

>Seems to be not significant as pick_up

#Scoring
##We can do some sort of scoring that indecate that trip duration is normal or not
```{r}
train_geo_data$category <- 
  case_when(log(train_geo_data$median_duration) <= 2.5 ~ "low", 
          between(log(train_geo_data$median_duration), 2.5, 5) ~ "medium_low",
         between(log(train_geo_data$median_duration), 5, 7.5) ~ "medium_high",
         log(train_geo_data$median_duration) > 7.5 ~ "high")

```


```{r}
train_geo_data %>% 
  group_by(category) %>% 
  summarise(number_of_locations = n()) %>% 
  ungroup %>%
  ggplot(aes(x=category, y=number_of_locations)) + 
  geom_bar(stat="identity", fill="#E41A1C") +
  scale_y_continuous() +
  labs(x="Category", y="No of pick up Locations ") +
  theme_minimal()
```

>Most of pickup locations fall in medium_high trip duration category

##Prepare model for linea regression

###1- Transform Longitude and Latitude using PCA
```{r}
library(stats)
pca_pickup <- princomp(train[c("pickup_longitude","pickup_latitude")], cor = TRUE)
pca_dropoff <- princomp(train[c("dropoff_longitude","dropoff_latitude")], cor = TRUE)

train$pick_lon_pc1 <- pca_pickup[["scores"]][,"Comp.1"]
train$pick_lon_pc2 <- pca_pickup[["scores"]][,"Comp.2"]

train$drop_lon_pc1 <- pca_dropoff[["scores"]][,"Comp.1"]
train$drop_lon_pc2 <- pca_dropoff[["scores"]][,"Comp.2"]
```


###2-Convert categorical variables
```{r}
table(train$pickup_weekend)
train$pickup_weekend <- ifelse(train$pickup_weekend=="Weekend",1,0)

train$vendor_id <- ifelse(train$vendor_id==2,1,0)
train$store_and_fwd_flag <- ifelse(train$store_and_fwd_flag=="N",1,0)
```

##Select features for the model
```{r}
my_features <- train %>% 
  select(
    vendor_id, passenger_count, store_and_fwd_flag, trip_duration, pickup_month,
    pickup_weekdays, pickup_hour, pickup_minute, total_distance, number_of_steps, pick_lon_pc1,
    pick_lon_pc2, drop_lon_pc1, drop_lon_pc2, peak_hour
  )
```

##Fit linear regression model
```{r}
linearmodel <- Regression_df %>% 
  select(-id, -pickup_longitude, -pickup_latitude, -dropoff_latitude, -dropoff_longitude,-starting_street, -end_street, -street_for_each_step, -distance_Haversine, -total_travel_time, -vendor_id, -store_and_fwd_flag, -pickup_minute,-number_of_steps) %>% 
  lm(formula = trip_duration ~.) %>% 
  summary()
```

##Calculate RMSE
```{r}
RSS <- c(crossprod(linearmodel$residuals))
MSE <- RSS / length(linearmodel$residuals)
10^(sqrt(MSE))
```

#Use our test data to check for the erros

##Make same peparations for test data
```{r}
names(test)
```

```{r}
f_routes <- fastest_routes_test %>% 
  select(id, starting_street, end_street, street_for_each_step, total_distance)

test <- left_join(test, f_routes, by="id")
names(test)
```

```{r}
test$total_distance <- log10(test$total_distance)
test$vendor_id <- ifelse(test$vendor_id==2,1,0)

test$pickup_hour <- hour(test$pickup_datetime)
test$pickup_week <- week(test$pickup_datetime)
test$pickup_month <- month(test$pickup_datetime)
test$pickup_weekdays <- weekdays(test$pickup_datetime)
test$pickup_weekend <- ifelse(test$pickup_weekdays=="Thursday"|test$pickup_weekdays=="Friday","Weekend","not-Weekend")

test = as.data.table(test)

test[,pickup_datetime:=as.Date(pickup_datetime)]


test[,":="(
   pickup_yday=yday(pickup_datetime)
  ,pickup_mday=mday(pickup_datetime)
)]
test <- as.data.frame(test)

library(stats)
pca_pickup <- princomp(test[c("pickup_longitude","pickup_latitude")], cor = TRUE)
pca_dropoff <- princomp(test[c("dropoff_longitude","dropoff_latitude")], cor = TRUE)

test$pick_lon_pc1 <- pca_pickup[["scores"]][,"Comp.1"]
test$pick_lon_pc2 <- pca_pickup[["scores"]][,"Comp.2"]

test$drop_lon_pc1 <- pca_dropoff[["scores"]][,"Comp.1"]
test$drop_lon_pc2 <- pca_dropoff[["scores"]][,"Comp.2"]
```


###2-Convert categorical variables
```{r}
table(train$pickup_weekend)
train$pickup_weekend <- ifelse(train$pickup_weekend=="Weekend",1,0)

test_featues <- test %>% 
  select(-id, -pickup_longitude, -pickup_latitude, -dropoff_latitude, -dropoff_longitude,-starting_street, -end_street, -street_for_each_step)
```


```{r echo=FALSE}
ypred <- Regression_df %>% 
  select(-id, -pickup_longitude, -pickup_latitude, -dropoff_latitude, -dropoff_longitude,-starting_street, -end_street, -street_for_each_step, -distance_Haversine, -total_travel_time, -vendor_id, -store_and_fwd_flag, -pickup_minute,-number_of_steps) %>% 
  lm(formula = trip_duration ~.) %>% predict(data=test_featues)
```

###MSE
```{r echo=FALSE}
ypred <- as.data.frame(ypred)
RMSE <- sqrt(mean((ypred - train$trip_duration)^2))
10^(RMSE)
```

>This is very larg error!

##XGBoost Algorithm
```{r}
library(caret)
library(xgboost)
x <- Regression_df[1:100000,]
samplesize = 0.70 * nrow(x)
set.seed(80)
index = sample( seq_len ( nrow ( x) ), size =samplesize )

train_n <- x[index , ]
test_n <- x[-index , ]
TrainControl <- trainControl( method = "repeatedcv", number = 5, repeats = 2)
model<- train(trip_duration ~ ., data = train_n, method = "xgbLinear", trControl = TrainControl,verbose = FALSE)
model
predicted <- predict(model, test_n)
c <- predicted - test_n$trip_duration
v = sqrt(mean(c^2)) ##8 minutes
10^(0.365)

t <- Regression_df %>% 
  select(-id, -pickup_longitude, -pickup_latitude, -dropoff_latitude, -dropoff_longitude,-starting_street, -end_street, -street_for_each_step, -distance_Haversine, -total_travel_time)
t2 <- t[1:30000,]
predicted <- predict(model, t2)

v <- sqrt(mean((predicted - t2$trip_duration)^2)) ##8 minutes


```

